Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-08-03 bmc_template/manuscript.tex
fig caps and ethics statement?

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}         
   email={lim88@mcmaster.ca}   % email address
]{\inits{ML}\fnm{Michael} \snm{Li}}
\author[
   addressref={aff1,aff2},
   email={bolker@mcmaster.ca}
]{\inits{BMB}\fnm{Benjamin M.} \snm{Bolker}}

@@

Clustering time-series data into discrete groups can improve 
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers. More generally, we point out that model misspecification
(omitting important sources of variation) can lead to overfitting,
and as a corollary that incorporating previously neglected
structure in statistical models can allow more
accurate assessment of appropriate model complexity.

\parttitle{Background} %if any
Text for this section.

\parttitle{Results} %if any
Text for this section.

\parttitle{Conclusion} %if any
Text for this section.

@@

\begin{equation}
\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}

@@

@@

\section*{Competing interests}
  The authors declare that they have no competing interests.

\section*{Author's contributions}
    Text for this section \ldots

\section*{Acknowledgements}
  Text for this section \ldots
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}

>>>>

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}         
   email={lim88@mcmaster.ca}   % email address
]
{\inits{ML}\fnm{Michael} \snm{Li}}
\author[
   addressref={aff1,aff2},
   email={bolker@mcmaster.ca}
]{\inits{BMB}\fnm{Benjamin M.} \snm{Bolker}}

@@

% More generally, we point out that model misspecification
% (omitting important sources of variation) can lead to overfitting,
% and as a corollary that incorporating previously neglected
% structure in statistical models can allow more
% accurate assessment of appropriate model complexity.

\parttitle{Background} %if any
Clustering time-series data into discrete groups can improve 
prediction and provide insight into the nature of underlying,
unobservable states of the system. However, temporal variation in the
rates at which individuals move between groups can obscure
such signals. We use finite mixture and hidden Markov models (HMMs), two
standard clustering techniques, to model high-resolution hourly
movement data from Florida panthers (\emph{Puma concolor coryi}).
Allowing for temporal heterogeneity in transition probabilities, a straightforward 
but little-used extension of the standard HMM framework, 
resolves some shortcomings of current
models and clarifies the behavioural patterns of
panthers.

\parttitle{Results} %if any
Simulations and Florida panthers data showed model misspecification (omitting important 
sources of variation) can lead to overfitting and over-estimating number of behavioural states. Models incorporating temporal heterogeneity have lower number of states with slightly higher variation in short movement states, and able to make out of sample predictions that captures observed diurnal and autocorrelation patterns exhibited by Florida panthers.

\parttitle{Conclusion} %if any
Incorportating temporal heterogeneity reduce the selected number of behavioural states closer 
to a biologically interpretable level, improved goodness of fit and predictability. Our 
suggest that incorporating previously neglected structure in statistical models can
allow more accurate assessment of appropriate model complexity.

@@

\begin{equation}
\textrm{Number of Free Parameter}=\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}

@@

\section{Declarations}

@@

\section*{Acknowledgements}
  Florida Wildlife and Fisheries?

\section{Data accessibility}
Hourly step lengths and turning angles of male and female Florida pan- thers available at: http://ufdc.ufl.edu//IR00004241/00001.

\section*{Author's contributions}
    Equally contributed.

\section*{Competing interests}
  The authors declare that they have no competing interests.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                  The Bibliography                       %%
%%                                                         %%
%%  Bmc_mathpys.bst  will be used to                       %%
%%  create a .BBL file for submission.                     %%
%%  After submission of the .TEX file,                     %%
%%  you will be prompted to submit your .BBL file.         %%
%%                                                         %%
%%                                                         %%
%%  Note that the displayed Bibliography will not          %%
%%  necessarily be rendered by Latex exactly as specified  %%
%%  in the online Instructions for Authors.                %%
%%                                                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section*{Figures}
  \begin{figure}[h!]
  \includegraphics[width=\maxwidth]{figure/sim_results-1}
  \caption{\csentence{Sample figure title.}
      A short description of the figure content
      should go here.}
      \end{figure}
